---
title: "PCA_of_VH_EN_Treatment"
author: "Laura Hover"
date: "11/7/2020"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Principal component analysis of vehicle vs. entrectinib treated samples.  LogCPM values were generated in the script: BulkRNAseq_DiffExpAnalysis.Rmd.  

```{r}
library(tidyverse)
library(FactoMineR)
library(ggplot2)
library(stringr)
library(RColorBrewer)
library(plotly)
```

Read in logCPM values:
```{r}
logCPM<-read.delim("VH_EN_logCPM.txt", header=T, sep="\t", stringsAsFactors = FALSE)
head(logCPM)
```

Create table of ensemblID and gene symbol to match up throughout analysis
```{r}
Genes<-logCPM[, 1:2]
```



PCA with logCPM expression
```{r}
#transpose matrix so observations are rows and variables are columns:
row.names(logCPM)<-logCPM$gene_id
logCPM<-logCPM[,-c(1:2)]
dat<-data.frame(t(logCPM))

pc <- prcomp(dat, scale=FALSE, center=TRUE) #default - centers and scales

summary(pc)
#Can see from summary PC1 explains 57% of the variance, PC2 explains 11%, PC3 9%

screeplot(pc)
```


#Look at PCA and top loadings for PC1 and PC2
```{r}
# loadings
#head(pc$rotation) #ensembl IDs

#x=the coordinates of the individuals (observations) on the principal components.
PC1 <- as.matrix(x=pc$x[,1]) 
PC2 <- as.matrix(pc$x[,2])
Samples<-colnames(logCPM)
Treat<-c("VH", "VH", "EN", "EN", "VH", "EN", "VH", "EN", "VH", "VH", "EN")
Gender<-c("F", "M", "F", "F", "M", "F", "F", "F", "M", "M", "F")

PC <- data.frame(Samples, Treat, Gender, PC1, PC2)
ggplot(PC, aes(PC1, PC2, color=Samples)) + 
  geom_text(aes(label = Samples), size = 5) +
  xlab("PC1") + 
  ylab("PC2") + 
  ggtitle("First Two Principal Components")+
  theme_classic()
```

Top loadings for PC1 and 2
```{r}
PC1_Loadings<-pc$rotation[,1]
PC1_Loadings<-sort(PC1_Loadings, decreasing=TRUE)
PC1_Loadings<-data.frame(PC1_Loadings)
PC1_Loadings$gene_id<-row.names(PC1_Loadings)
PC1_Loadings<-left_join(PC1_Loadings, Genes, by="gene_id")
#Many of the top loadings for PC1 include cell cycle genes like: Ccnd1, Cdk1, Top2a, Ccnb1, Rasl11a, Cdkn2a


PC2_Loadings<-sort(pc$rotation[,2], decreasing=TRUE)
PC2_Loadings<-data.frame(PC2_Loadings)
PC2_Loadings$gene_id<-row.names(PC2_Loadings)
PC2_Loadings<-left_join(PC2_Loadings, Genes, by="gene_id")
#Neuronal genes: Neurod6, Neurod1, Emx1, Foxg1, Ddn
```


```{r}
par(mfrow=c(2,1))
barplot(PC1_Loadings$PC1_Loadings[1:25],cex.names=.6,main="PC 1 Loadings", las=2, names.arg = PC1_Loadings$gene_symbol[1:25]) #loadings for variables in linear combination for PC1
barplot(PC2_Loadings$PC2_Loadings[1:25],cex.names=.6,main="PC 2 Loadings", las=2, names.arg=PC2_Loadings$gene_symbol[1:25])

par(mfrow=c(2,1))
barplot(PC1_Loadings$PC1_Loadings[16792:16817],cex.names=.6,main="PC 1 Loadings", las=2, names.arg=PC1_Loadings$gene_symbol[16792:16817]) #loadings for variables in linear combination for PC1
barplot(PC2_Loadings$PC2_Loadings[16792:16817],cex.names=.6,main="PC 2 Loadings", las=2, names.arg = PC2_Loadings$gene_symbol[16792:16817])
```

PCs 2 and 3, look at top loadings for PC3
```{r}
#x=the coordinates of the individuals (observations) on the principal components.
PC2 <- as.matrix(x=pc$x[,2]) 
PC3 <- as.matrix(pc$x[,3])

PC <- data.frame(Samples, Treat, PC2, PC3)
ggplot(PC, aes(PC2, PC3, color=Treat)) + 
  geom_point(aes(label = Samples), size = 5) +
  xlab("PC2") + 
  ylab("PC3") + 
  ggtitle("Principal Components 2 and 3")+
  theme_classic()
```

PC3 Loadings:
```{r}
PC3_Loadings<-sort(pc$rotation[,3], decreasing=TRUE)
head(PC3_Loadings)

barplot(PC3_Loadings[1:25],cex.names=.6,main="PC 3 Loadings", las=2) #loadings for variables in linear combination for PC1
```

#3D PCA with top 1000 most variable genes
Select top 1000 most variable genes
Read in data if this is the starting point on the script
```{r}
#Read in logCPM file:
logCPM<-read.delim("VH_EN_logCPM.txt", sep = "\t", stringsAsFactors = FALSE)

genes<-logCPM$gene_id
logCPM<-logCPM[,-c(1:2)]
row.names(logCPM)<-genes


#in logCPM, rows are genes, columns are samples, find most variable across the columns (samples)
mostvar <- apply(logCPM, 1, var)
#This is now a vector of the variance for each gene

#Next steps: sort genes by variance, taking the top 1000, keeping the gene IDs, merge with logCPM to get expression of the top 1000 most variable genes
mostvar<-sort(mostvar, decreasing = TRUE)

#Keep top 1000 most variable genes
mostvar<-mostvar[1:1000]

#Keep only the most variable genes in our logCPM table:
var<-logCPM[mostvar,]
```

Change format so the rownames are the sample IDs and the columns contain expression values only
```{r}
## tidyverse manipulation of the data to change the format
var<-data.frame(t(var))
```


run PCA on samples and generates eigenvectors
```{r}
pca<-FactoMineR::PCA(var, scale.unit=FALSE, ncp=5, graph=TRUE)
ev<-pca$eig
```

generates strings with the percentage of the variation assigned to each of the first five PCs
```{r}
PC1.per<-paste("PC1 (", round(ev[1,2], 1), "%)", sep = "")
PC2.per<-paste("PC2 (", round(ev[2,2], 1), "%)", sep = "")
PC3.per<-paste("PC3 (", round(ev[3,2], 1), "%)", sep = "")
PC4.per<-paste("PC3 (", round(ev[4,2], 1), "%)", sep = "")
PC5.per<-paste("PC3 (", round(ev[5,2], 1), "%)", sep = "")
```

extract the coordinates for all of our samples on the 1st 5 PC dimensions
```{r}
Sample_coord<-round(pca$ind$coord,4)
Sample_coord<-data.frame(Sample_coord)
Sample_coord<-rename(Sample_coord, PC1=Dim.1, PC2=Dim.2, PC3=Dim.3, PC4=Dim.4, PC5=Dim.5)
```

set up the groups for coloring the plot based on the meta data
```{r}
meta<-data.frame(row.names(var))
meta$Treat<-c(str_sub(meta$row.names.var.,1,2))  # Make new column with just EN or VH from each sample name
names(meta)<-c("Samples", "Treat")

#Set color scheme, colors from RColorBrewer
cols <- brewer.pal(4, "Paired")
set1<-c("#1F78B4", "#33A02C")
```

## interactive 3D plot
Use a negative on the Z axis to help visualize the clusters/samples
```{r}
p <- plot_ly(Sample_coord, x = ~-PC1, y = ~PC2, z = ~-PC3, type = "scatter3d",
color = meta$Treat, colors = set1,
marker=list(size=20),
#symbol = meta$Treat, symbols = c("0", "1"), # if you want symbols by region remove # from start of this line
text = meta$Samples, hoverinfo = "text") %>%
layout(title = "top 1000", scene = list(
xaxis = list(title = PC1.per, titlefont=list(size=20), gridcolor=toRGB("gray50"),
  gridwidth = 4, tickfont=list(size=16)),
yaxis = list(title = PC2.per, titlefont=list(size=20), gridcolor=toRGB("gray50"),
  gridwidth = 4, tickfont=list(size=16)),
zaxis = list(title = PC3.per, titlefont=list(size=20), gridcolor=toRGB("gray50"),
  gridwidth = 4, tickfont=list(size=16))
))

p

```

